Learning Actionable Analytics (with applications for reducing defects and reducing runtimes)

### 2015-05-06 how.tex
Updates from ShareLaTeX

 \documentclass[conference]{IEEEtran}
\usepackage{subfig}
\usepackage{wrapfig}
 \usepackage{amsmath}
 \usepackage{url}
 \usepackage{pifont}
 \usepackage{times}
\usepackage{rotating}
%\usepackage{balance} 
\usepackage{color, colortbl}
\usepackage{graphicx}
\usepackage{algorithmicx}

\usepackage[running]{lineno}
\usepackage{program}
\usepackage{cite}
\usepackage{alltt}
\newcommand{\eq}[1]{Equation~\ref{eq:#1}}
\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\be}{\begin{enumerate}}
\newcommand{\ee}{\end{enumerate}}
\newcommand{\tion}[1]{\textsection\ref{sec:#1}}
\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
\definecolor{lightgray}{gray}{0.975}
\usepackage{fancyvrb}
\usepackage{stfloats}
\usepackage{multirow}
\usepackage{listings}
\usepackage{amsmath} 
\DeclareMathOperator*{\argmin}{arg\,min} 
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\definecolor{Code}{rgb}{0,0,0}
\definecolor{Decorators}{rgb}{0.5,0.5,0.5}
\definecolor{Numbers}{rgb}{0.5,0,0}
\definecolor{MatchingBrackets}{rgb}{0.25,0.5,0.5}
\definecolor{Keywords}{rgb}{0,0,1}
\definecolor{self}{rgb}{0,0,0}
\definecolor{Strings}{rgb}{0,0.63,0}
\definecolor{Comments}{rgb}{0,0.63,1}
\definecolor{Backquotes}{rgb}{0,0,0}
\definecolor{Classname}{rgb}{0,0,0}
\definecolor{FunctionName}{rgb}{0,0,0}
\definecolor{Operators}{rgb}{0,0,0}
\definecolor{Background}{rgb}{1,1,1}

\lstnewenvironment{python}[1][]{
\lstset{
mathescape,
numbers=#1,
numberstyle=\footnotesize,
numbersep=0.5em,
xleftmargin=0em,
framextopmargin=2em,
framexbottommargin=2em,
showspaces=false,
showtabs=false,
showstringspaces=false,
tabsize=2,
% Basic
basicstyle=\ttfamily\small\setstretch{0.8},
backgroundcolor=\color{Background},
language=Python,
% Comments
commentstyle=\color{Comments}\slshape,
% Strings
stringstyle=\color{Strings},
morecomment=[s][\color{Strings}]{"""}{"""},
morecomment=[s][\color{Strings}]{'''}{'''},
% keywords
morekeywords={[1]import,from,class,def,for,while,if,is,in,elif,else,not,and,or,print,break,continue,return,True,False,None,access,as,,del,except,exec,finally,global,import,lambda,pass,print,raise,try,assert, dot, norm, zip, sorted},
keywordstyle={[1]\color{Code}\bfseries},
% additional keywords
morekeywords={[3]fastmap, project, furthest, split, WHERE,clusterer, getContours, envied, fWeight, nearestContour, projection, mutate, HERE, knn},
keywordstyle={[3]\color{Keywords}\bfseries},
morekeywords={[2]@invari},
keywordstyle={[2]\color{Decorators}\slshape},
emph={self},
emphstyle={\color{self}\slshape},
%
}}{}

@@

  \begin{abstract}
 As business users demand more insightful
 analytics, data scientists need to change
 their tools. Instead of merely predicting 
 some result, they also need tools that generate ``policy advice'';
 i.e. suggestions on  how to change values in order to
 improve on some predicted value.

 This paper proposes one such policy
 generator. ``HOW'' is a near-linear-time
 tool for proposing changes to independent
 variables in order to improve on 
 the predictions of the dependent variables. Unlike other approaches
 for learning optimizations to software does not require
 an underlying of the domain (rather, HOW automatically
 learns that model as part of its operation).

 This paper uses  HOW to learn methods
 to reduce defects and decrease program runtime.
 For the test data of this paper, the improvements generated by HERE can reduce
 defect counts as well as runtimes by up to
 50\% of their original values.
  \end{abstract}
  \begin{IEEEkeywords}
Instance-based reasoning, defect prediction, random forest, optimization.
  \end{IEEEkeywords}

\section{Introduction}
Business  users are now demanding that data mining tools
be augmented with tools to support  business-level
interpretation of that data. For example,
at a  panel on software analytics at ICSE'12,
industrial practitioners lamented the state of the art in data mining
and software engineering~\cite{menzies12a}. Panelists commented that
``prediction is all well and good, but what about decision
making?''. That is, these panelists are more interested in the interpretations
that follow the mining, rather than just  the mining. For example:
\bi
\item Instead of just accepting  {\em predictions} on how many 
 software defects
to expect,  business users might now demand a {\em plan} to
reduce the likelihood of those defects;
\item Instead of just accepting {\em predictions} on the runtime
time of their software, business users might now demand
a {\em plan} to reduce that runtime.
\ei
In these two points, we say that a ``plan'' is a  set of proposed changes in order to better achieve some goal. 
Generating {\em plans} is a  very different task the standard
data mining tasks of {\em predicting} some discrete or numeric class.
Predictors
and classifiers only tell us what {\em is} while planners 
  tell us {\em what to change}. For the kind of business user
discussed above, planners are the preferred to mere prediction systems
since, as one gruff user once shouted at us:
\begin{quote}
{\em ``Don't tell me what \underline{is};
tell me what to \underline{change}.''}
\end{quote}
In order to satisfy the demands of this new kind of business user, we  study the generation of {\em plans} in the context
of software analytics.  
This paper presents
HOW, a novel  {\em instance-based planner} that runs in near-linear time.
In the experiments shown below, we see that HOW's plans can
significantly reduce the expected
value of:
\bi
\item The software defects found in  twelve object-oriented JAVA systems explored in 2010 by Jureczko et al.~\cite{jureczko10};
\item The runtime of the six software systems whose configurations
are explored in 2013 by Seigmund et al.~\cite{sven12};
\ei
In summary, this paper makes six contributions that are significantly different to
prior work:
\be
\item Many software analytics papers use data to build {\em predictors} for (e.g.) defects and runtimes.
However, aside from work by the authors of this paper,
none of those papers then go further to use that data to build {\em planners}.
\item This paper offers a general framework for augmenting a prediction system such that
it becomes a planning system;
\item This paper offers a  test that recognizes  when we should {\em not}  plan: in summary, our kind of planning is
not recommended in domains that  generate poorly performing predictors;
\item This paper offers a general verification rig for assessing the value of plans generated within this rig.
\item The HOW  instance-based planner of this paper has not been presented before.
\item The experiments of this paper show that HOW planner is useful for real-world tasks;
e.g. reducing the
expected values of the Jureczko software defects and the Seigmund software runtimes.
\ee
The rest of this paper is structured as follows.
HOW's instance-based reasoning  makes its conclusions by
studying the space of examples. As discussed in \tion{mm},
this is a very different approach to
{\em model-based planners} that use the model
as sub-routine within a Monte Carlo simulation~\cite{me07f} or the evolutionary 
programming methods explored by the search-based software
engineering community~\cite{krall14,harman12dec}.

@@

 \subsection{Motivation: Why ``HOW''?}
HOW was inspired by some recent work by Krall and Menzies and Davis on 
the GALE optimizer~\cite{krall14}. A well-known effect in machine learning (and, in computer vision
and compression~\cite{levina04}) is that the $n$ dimensions of a data set can be represented by a  much
lower-dimensional approximation. GALE finds those $m << n$ dimensions then mutates examples
towards the better end of each dimension $m_i$. Since it explores $m \ll n$ dimensions,
GALE terminates much faster than standard optimiser (e.g. for one large model, 4 minutes versus 7 hours~\cite{krall14}).
Better yet, GALE's rapdily generated  solutions are competitive (or better) than those found by 
slower optimizers.

A limitation with GALE is that it makes extensive use of
an underlying domain model to asses generations of newly mutated instances.  This is an issue since
in some domains, no such model exists or is readily accessible. For example, this whole paper
began when, after briefing a business client on GALE, they asked:
\begin{quote}
{\em ``But what if I do not have a model of my business?''}
\end{quote}
This is a legitimate concern.
Building and certifying a model  can  takes  long time, especially in software engineering. For  example, in previous work~\cite{me07f} we used models
developed by Boehm's group at the University of Southern California.
Some of those models took decades to develop and mature (from 1981~\cite{boehm81} to 2000~\cite{boehm00b}).  This is not an isolated example. One of us (Menzies) teaches a graduate class in automated model-based software engineering. 
Students in that class are asked to find (or build) a non-trivial model of
some software process (where ``non-trivial'' means something that processes
a  model at least an order of magnitude more complex than Brooks' Law). 
While the  number of such models has been growing in recent years\footnote{See the
catalogue at goo.gl/nv2AVK.}, 
the total number remains very small-- which is an observation consistent with
SE models typically require  much effort to build and certify.

Also, even when there is an existing model, instance-based reasoning can be useful.
Models need maintenance lest they become out-dated. Elsewhere, we have described our
extensions to the USC models to enable reasoning about agile software developments. 
It took many months to implement and certify those extensions~\cite{me09i,me09j}.
The problem of model maintenance is another
motivation to look for alternate methods that can be automateically updated whenever new data arrives.

Finally, sometimes  it is easier to access instances of a model's behaviour than the model
itself. For example, in prior work with Martin  Feather from the Jet Propulsion 
Laboratory~\cite{fea02a},  our research partner could not share a
propriety model from within the NASA firewalls. However, they could share 
logs of the input/output of that model.

Note that this lack of access to a workable model is not just fatal for GALE, but also for every
other model-based optimization algorithm such as simulated
annealling or evolutuionary methods such as   
NSGA-II, NSGA-III, SPEA2, IBEA, PSO, DE, MOEA/D, etc.~\cite{deb00a,zit02,zit04,%
deb14,Cui2005a,storn97,zhang07:TEC}. Accordingly, this paper seeks
an optimization methods that avoids the problem of models that are missing, slow
to build, hard to maintain, or difficult to access.

\subsection{Methods: How to Build ``HOW''?}
HOW is a variant of GALE that does not need a underlying domain model.

When designing HOW, our premises were:
\bi
\item
This is no domain model; (b)~we have some
labelled data and (b)~we seek minimal
disrupttion
HOW divide training into regions where the training
data has similar properties. Each region is then examined and divided into ``west'' and ``east''
where the region's ``west'' end has examples with worst performance stores than the east
end (mnemoc: west starts with ``w'' and is short for ``worst' and east starts with ``e;;]which is shirt
for excellent).so the ``e'' in

\be
\item Take  a population $P$ of examples
\item Separate all $p \in P$ examples into their $n$ independent variables and 
their $o$ scores;
\item Find $m \ll n$ lower dimensional approximations for the $n$ independent variables;
\item Look for the worst and best ends of each $m$  dimension (where the better ends
have better objective scores);
\item Find the $q \subseteq P$ subset of the population with ``better'' objective scores;
\item Generate the next population $P'$ by adjust the independent variables
of the $q$  examples
by a random amount (in the direction of  the  better ends of each dimension).
\item Apply some model to discover all the objective scores of $P'$
\item Set $P=P'$ and go to step 1
\ee

PCA explores the $e$
eigenvectors of the correlation matrix between $d$
variables~\cite{pearson1901}.  Usually,
only a few $e \ll d$ eigenvectors are required
since multiple inter-related variables $i,j,k \subseteq d$ can be represented
by a single  eigenvector.
Also, noisy variables from $d$ are
ignored since they  do not contribute to the signal in the data.
Further,
redundant  variables  become (approximately) parallel lines
in $e$ space. For all such redundancies \mbox{$i,j \in d$}, we
can ignore $j$
since effects that change over $j$ also
change in the same way over $i$.

Algorithms like PDDP~\cite{boley98} use PCA to
recursively divide data into smaller regions. Each
step of that process requires an $O(N^2)$ analysis
of the data to compute the components~\cite{ilin10}. Worse, it requires an evaluation of every example.
Using WHERE, competitive results can be achieved
using an $O(2N)$ analysis.
WHERE uses a heuristic proposed by
Faloutsos and Lin~\cite{Faloutsos1995}
(which
Platt has
  shown computes a Nystr\"om
  approximation to the first component of
  PCA~\cite{platt05}).
In step one, WHERE inputs
$N$
examples $N_1,N_2,..$. Secondly, WHERE
picks any
point $N_i$ at random. Thirdly, WHERE
finds the point  {\em West}~$\in N$ that is
furthest\footnote{
 For this work, we use the standard Euclidean measure recommended for
instance-based reasoning by Aha et al.~\cite{aha91};
i.e. $\sqrt{\sum_i(x_i-y_i)^2}$ where $x_i,y_i$
are values normalized 0..1 for the range min..max.}
from $N_i$.
Finally, WHERE

$m \ll n$
dimension

\be
\item Clustering the training data 
\item 
(c) finding which end of the cluster has better performance scores.
(d) find
\ee

\section{Methodological Comments:\\ from ``Prediction'' to ``Planning''}
A major  problem with building a planning mechanism in software engineering is ``how to verify the generated
plans?''. For example,  in the experiments shown below,  HOW comments  on how to reduce
software defects by adjusting certain static code parameters of the code\footnote{Such parameters can be mnaually adjusted by programmers or (for large scale changes) automatically adjusted using code refactoring
tools, as shown by Binkely et al. in 2006~\cite{Binkley2006} and more recently by Mkaouer et al. in 2015~\cite{Mkaouer15}.}. How are we to verify that HOW's recommendations actually work?

Very few industrial or research groups have access
to the kinds of resources using by Bente et al. (evidence: in the size years since the
publication of that work, we know of only one of two similar studies). Hence, it
we wish to verify HOW's plans, then some other approach is required that is less
resource intensive.

\subsection{Proposed Framework}

We propose using prediction systems (built by data mining) as a verification
tool for planners. In domains with data, from which data miners can build
predictors of adequate performance, then those predictors can assess the value
of the changes proposed by HOW.
As shown in \fig{work}, 
project data is divided into two disjoint sets {\em train} and {\em test}
(so \mbox{{\em train} $\cup$ {\em test} $=\;\emptyset$}).
Next, from the train set, we build both a {\em planner} as well
as a {\em predictor}. Our general framework does not   commit to any particular  choice
of {\em planner} or {\em predictor}. For the purposes of this discussion 
our {\em planner}
will be HOW (discussed below)and our
our {\em predictors} will be the Random Forest and Random Forest
Regressor\footnote{Decision tree learners find the attribute whose value's
select for for sections of the data with least variation in the class
variable. Decision tree learners then recurse on each selection~\cite{breiman84}. A random forest is a set of decision trees, each built
from some random subset of the rows and columns. The conclusions
of the forest are a ``vote'' generated by all the trees~\cite{Breiman2001}. 
A standard Random Forest predicts for a discrete class variable while
a Random Forest Regressor generates numeric predictions.} taken from the SciKit
Learn toolkit~\cite{Pedregosa2012}.

Turning now to the {\em test} data, this is passed to the {\em predictor}
to generate a baseline {\em before} performance score.
Note that if our {\em predictors} fail to perform adequately on the test data,
then we cannot trust them to comment on our plans. Accordingly,
if that performance is unsatisfactory, we abort.
Else, we (1)~apply the {\em planner} to alter the {\em test} data;
then (2)~apply the {\em predictor} to the altered data {\em test'};
then (3)~return the performance delta.

@@

\subsection{Are These Plans ``Causal''?}
With one additional stipulation,   this framework can be used to check if the learned
plans can be said to {\em cause} a change in the domain being explored.

In the literature, there are two main definitions of causality: Hill's H-causality~\cite{Hill1965}
and Granger's G-casualty~\cite{granger80}. 
Of the two, Hill's definition is the most strict. An effect
is H-causal if it satisfies nine criteria including
{\em  plausibility};
{\em strength} of the effect;
{\em 
specificity}, i.e. locality of the effect
to specialized groups;
{\em
support} by experimental evidence;
{\em 
coherence} between cause-effects noted
in the field and in laboratory samples;
{\em
temporally} i.e. the effect has to follow the cause;
{\em 
analogy}, i.e.  similar causes have similar effects;
{\em
biological gradient}, i.e.  more exposure to the cause leads to more of an
effect; and
{\em
consistency}, i.e.  the same cause-effect pairings have been seen by different persons
in different places.

It is hard to demonstrate H-causality in software engineering.
For example, Hill's criteria emphasises repeatability and laboratory studies-- both of which are
hard to perform in software engineering (e.g. see the continued debate about
the value of using undergraduate students to surrogates for industrial
practitioners in software engineering research~\cite{Carver2003}). Also, Hill's requirement
for consistency is far easier to achieve in studies of small closed systems (e.g. bacteria in a petri dish)
than in large socio-technical systems (e.g. open source projects). In the latter,
effects are more {\em probabilistic} than {\em categorical}-- by which we mean that 
some cause tends to cause some effect at some probability $p$ less that one.
This in turn means that at some probability $q$ (where $0 \le q < p \le 1$)
that we will find examples of {\em inconsistency} where some cause does {\em not} lead
to some effect (thus violating one of Hill's criteria for demonstrating causality).

>>>>

 \documentclass[conference]{IEEEtran}
\usepackage{subfig}
\usepackage{wrapfig}
 \usepackage{amsmath}
 \usepackage{url}
 \usepackage{pifont}
 %\usepackage{times}
\usepackage{rotating}
%\usepackage{balance} 
\usepackage{color, colortbl}
\usepackage{graphicx}
\usepackage{algorithmicx}

\usepackage[running]{lineno}
\usepackage{program}
\usepackage{cite}
\usepackage{alltt}
\newcommand{\eq}[1]{Equation~\ref{eq:#1}}
\newcommand{\bi}{\begin{itemize}}
\newcommand{\ei}{\end{itemize}}
\newcommand{\be}{\begin{enumerate}}
\newcommand{\ee}{\end{enumerate}}
\newcommand{\tion}[1]{\textsection\ref{sec:#1}}
\newcommand{\fig}[1]{Figure~\ref{fig:#1}}
\definecolor{lightgray}{gray}{0.975}
\usepackage{fancyvrb}
\usepackage{stfloats}
\usepackage{multirow}
\usepackage{listings}
\usepackage{amsmath} 
\usepackage[caption=false,font=footnotesize]{subfig}
\DeclareMathOperator*{\argmin}{arg\,min} 
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\definecolor{Code}{rgb}{0,0,0}
\definecolor{Decorators}{rgb}{0.5,0.5,0.5}
\definecolor{Numbers}{rgb}{0.5,0,0}
\definecolor{MatchingBrackets}{rgb}{0.25,0.5,0.5}
\definecolor{Keywords}{rgb}{0,0,1}
\definecolor{self}{rgb}{0,0,0}
\definecolor{Strings}{rgb}{0,0.63,0}
\definecolor{Comments}{rgb}{0,0.63,1}
\definecolor{Comments}{rgb}{0.5,0.5,0.5}
\definecolor{Backquotes}{rgb}{0,0,0}
\definecolor{Classname}{rgb}{0,0,0}
\definecolor{FunctionName}{rgb}{0,0,0}
\definecolor{Operators}{rgb}{0,0,0}
\definecolor{Background}{rgb}{1,1,1}

\lstnewenvironment{python}[1][]{
\lstset{
mathescape,
numbers=none,
numberstyle=\scriptsize,
numbersep=0.5em,
xleftmargin=0em,
framextopmargin=2em,
framexbottommargin=2em,
showspaces=false,
showtabs=false,
showstringspaces=false,
tabsize=2,
% Basic
basicstyle=\ttfamily\scriptsize\setstretch{0.8},
backgroundcolor=\color{Background},
language=Python,
% Comments
commentstyle=\color{Comments}\slshape,
% Strings
stringstyle=\color{Strings},
morecomment=[s][\color{Strings}]{"""}{"""},
morecomment=[s][\color{Strings}]{'''}{'''},
% keywords
morekeywords={[1]import,from,class,def,for,while,if,is,in,elif,else,not,and,or,print,break,continue,return,True,False,None,access,as,,del,except,exec,finally,global,import,lambda,pass,print,raise,try,assert, dot, norm, zip, sorted},
keywordstyle={[1]\color{Code}\bfseries},
% additional keywords
morekeywords={[3]fastmap, project, furthest, split, WHERE,clusterer, getContours, envied, fWeight, nearestContour, projection, mutate, HERE, knn},
keywordstyle={[3]\color{Keywords}\bfseries},
morekeywords={[2]@invari},
keywordstyle={[2]\color{Decorators}\slshape},
emph={self},
emphstyle={\color{self}\slshape},
%
}}{}

@@

  \begin{abstract}
 As business users demand more insightful
 analytics, data scientists need to change
 their tools. Instead of merely predicting 
 some result, they also need tools that generate ``plans'';
 i.e. specific suggestions on  how to change values in order to
 improve on some predicted value.

 This paper proposes one such planner. ``HOW'' is a 
 tool for proposing changes to independent
 variables in order to improve on 
 the predictions of the dependent variables. Unlike other approaches
 for learning optimizations to software, HOW does not require
 an underlying of the domain. Also, HOW's plans
 are designed to result in {\em minimal displacement}
 to current practice.

 This paper uses  HOW to learn methods
 to reduce defects and decrease program runtime.
 For the test data of this paper, the improvements generated by HERE can reduce
 defect counts as well as runtimes by up to
 50\% of their original values.
  \end{abstract}
  \begin{IEEEkeywords}
Defect prediction, configuration, prediction, planning, instance-based reasoning.
  \end{IEEEkeywords}

\section{Introduction}
Business  users   now demand   data mining tools
that  support a  business-level
interpretation of that data. For example,
at a  panel on software analytics at ICSE'12,
industrial practitioners lamented the state of the art in data mining
and software engineering~\cite{menzies12a}. Panelists commented that
``prediction is all well and good, but what about decision
making?''. That is, these panelists are more interested in the interpretations
and follow-up
that occurs {\em after} the mining, rather than just  the mining itself. For example:
\bi
\item Instead of just accepting  {\em predictions} on how many 
 software defects
to expect,  business users might now demand a {\em plan} to
reduce the likelihood of those defects;
\item Instead of just accepting {\em predictions} on the runtime
time of their software, business users might now demand
a {\em plan} to reduce that runtime.
\ei
In these two points, we say that a ``plan'' is a  set of proposed changes in order to better achieve some goal. 
Generating {\em plans} is a  very different task the standard
data mining tasks of {\em predicting} some discrete or numeric class.
Predictors
and classifiers only tell us what {\em is} while planners 
  tell us {\em what to change}. For the kind of business user
discussed above, planners are the preferred to mere prediction systems
since, as one gruff user once shouted at us:
\begin{quote}
{\em ``Don't tell me what \underline{is};
tell me what to \underline{change}.''}
\end{quote}
Generating plans is also differed to the recommender systems. 
discussed by Robillard and Walker~\cite{rob14}.  Recommendor systems
in software engineering
are less {\em directive} and more    {\em collaborative} than plan generators:
\bi
\item
Directive systems (such as those discussed in this paper) tell developers exactly what to do;
e.g. ``use these settings in a Makefile''. 
\item
On the other hand, the collaborative systems discussed by Robillard and Walker
are more suggestive guides that
help (e.g.) programmers focus on very small sets
with  very large libraries of code of documentation.
\ei
This paper presents HOW, a novel planner for learning actionable analytics.
In the experiments shown below, we see that HOW's plans can
significantly reduce the expected
value of:
\bi
\item The software defects found in  the object-oriented JAVA systems explored in 2010 by Jureczko et al.~\cite{jureczko10};
\item The runtime of the  software systems whose configurations
are explored in 2013 by Seigmund et al.~\cite{sven12};
\ei
In summary, this paper makes five contributions that are significantly different to
prior work:
\be
\item The HOW  instance-based planner of this paper has not been presented before.
\item The experiments of this paper show that HOW planner is useful for real-world tasks;
e.g. reducing the
expected values of the Jureczko software defects and the Seigmund software runtimes.
\item Apart from HOW, we  also offers a general framework for using  predictive system to build planners;
\item Further, we offer a  test to recognizes  when we should {\em not}  plan: specifically,
planning is
not recommended in domains with  poorly performing predictors;
\item Last, we  offer a general verification rig for assessing the value of plans generated within this rig.
\ee
The rest of this paper is structured as follows.
HOW's instance-based reasoning  makes its conclusions by
studying the space of examples. As discussed in \tion{mm},
this is a very different approach to
{\em model-based planners} that use the model
as sub-routine within a Monte Carlo simulation~\cite{me07f} or the evolutionary 
programming methods explored by the search-based software
engineering community~\cite{krall14,harman12dec}.

@@

 \subsection{Motivation: Why ``HOW''?}
HOW was inspired by two     business issues raised about the  GALE optimizer~\cite{krall14}. 
A well-known effect in machine learning (and, in computer vision
and compression~\cite{levina04}) is that the $n$ dimensions of a data set can be represented by a  much
lower-dimensional approximation. GALE finds those $m \ll n$ dimensions then mutates examples
towards the better end of each dimension $m_i$. Since it explores $m \ll n$ dimensions,
GALE terminates much faster than standard optimiser (e.g. for one large model, 4 minutes versus 7 hours~\cite{krall14}).
Better yet, GALE's rapdily generated  solutions are competitive (or better) than those found by 
slower optimizers.

When GALE executes, it searches for better examples of data
 by (a)~inputting an initial population of data; then (b)~making
 extensive mutation to that data. In practice, GALE makes
100s to 100,000s of mutations. Each of these mutations
is reassessed using some domain model.
For certain users,this approach raises some issues.  One such user commented:
\begin{quote}
ISSUE~\#1: {\em ``What if I only want the minimum useful change, not
the maximum possible?''}
\end{quote}
Note here that this business user is asking for solutions  that do not require extensive 
changes to their current  approaches.

Later in that discussion, another business user asked:
\begin{quote}
ISSUE~\#2: {\em ``But what if I do not have a domain model of my business?''}
\end{quote}
This also is a legitimate issue.
Building and certifying a model  can  takes  long time, especially in software engineering. For  example, in previous work~\cite{me07f} we used models
developed by Boehm's group at the University of Southern California.
Some of those models took decades to develop and mature (from 1981~\cite{boehm81} to 2000~\cite{boehm00b}).

Also, even when there is an existing model, they can require
constant  maintenance lest they become out-dated. Elsewhere, we have described our
extensions to the USC models to enable reasoning about agile software developments. 
It took many months to implement and certify those extensions~\cite{me09i,me09j}.
The problem of model maintenance is another
motivation to look for alternate methods that can be automatically updated whenever new data arrives.

Further, sometimes  it is easier to access instances of a model's behaviour than the model
itself. For example, in prior work with Martin  Feather from the Jet Propulsion 
Laboratory~\cite{fea02a},  our research partner could not share a
propriety model from within the NASA firewalls. However, they could share 
logs of the input/output of that model.

These two issues are not just problems for GALE, but also for every
other model-based optimization algorithm  that makes
extensive use of mutation  and domain models such as
NSGA-II, NSGA-III, SPEA2, IBEA, PSO, DE, MOEA/D, etc.~\cite{deb00a,zit02,zit04,%
deb14,Cui2005a,storn97,zhang07:TEC}. Accordingly, this paper seeks
an optimization methods that make minimal changes to instances,
while avoiding the problem of models that are missing, slow
to build, hard to maintain, or difficult to access.

\subsection{Methods: How to Build ``HOW''?}

A general model for {\em predictors} is as follows. 
During testing, each test instance gains a prediction 
by finding which region of the data it falls into.
During training, those regions are inferred from
divisions that ``chunk'' together regions with instances
with similar properties. These dividers can be categorical (as in the case of decision trees) or probabilistic (as in the
case of Naive Bayes) or learned via some distance minimization
process 
(as in the case of support vector machines) or generated
directly by some clustering algorithm.

\begin{figure}[t!]
\small
~\hrule~

{\bf \fig{where}a: top down clustering:}

Data can be rapidly and recursively divided   as follows.
Find   two   distance instances,  $X,Y$
by picking any instance $W$ at random, then setting $X$ to its most
distant instance, then setting $Y$ to the instance most distant from
$X$~\cite{fastmap}
(this requires only $O(2N)$ comparisons
of $N$ instances).
Next, project each instance $Z$
onto a line that  runs between $X,Y$ using the cosine
rule of \fig{where}b. After that,  split the data at the median $x$ value of all instances and
recurses on each half  (stopping when
one half has less  than $\sqrt{N}$ of the original population). See \fig{code1d} for further details.

~\hrule~

{\bf \fig{where}b: finding distances:}

In the \fig{where}a, to compute distances, we use
the Euclidean measure recommended for
instance-based reasoning by Aha et al.~\cite{aha91};
i.e. $\sqrt{\sum_iw_i(X_i-Y_i)^2}$ where $X_i,Y_i$
where values are  normalized 0..1 for the range min..max and 
$w_i$ is some importance weight given to attribute $i$.
Usually, $w_i=1$ but it can be adjusted using, say,
the feature weighting methods of \fig{where}d.

 ~\hrule~

{\bf \fig{where}c: 3 point geometry:}

Let   $X,Y$ be two points joined by  a line  length $c$.
A third point, $Z$, has distances  $a=dist(Z,X)$ and
$b=dist(Z,Y)$ to $X,Y$, respectively.
According to the cosine rule,   $Z$ projects onto  $\overline{XY}$
at distance $x=(a^2 + c^2 - b^2)/(2c)$ from $X$.
Further, according to Pythagoras' theorem, $Z$ stands at a distance
$y = \sqrt{a^2 - x^2}$ from the line $\overline{XY}$. See \fig{code1c} for further details.

~\hrule~

{\bf \fig{where}d: feature weighting:}

HERE's feature weighting algorithm
comes from the CART regression tree learner~\cite{Breiman1984}.
It sorts independent variables
 according to how well they can reduce the variance
of a  numeric objective.
If we split some column $i$ of independent numeric data  into $N=n_1 + n_2 + ..$ splits,
then the expected
value of the objective value of the split  is $w_i = \sum_i v_in_i/N$
where $v$ is the standard deviation of the objective.
A recursive procedure can  find those divisions $n_1,n_2,...$ by (a)~sorting a numeric column,
then split at $\argmin_i w_i$, then recursing into each half.  See \fig{code1d} for further details

~\hrule~

\caption{Sub-routines within HOW.}\label{fig:where}
\end{figure}

\input{code}

Those same regions can be used for instance-based 
{\em planning}-- but instead of leaving test instances in the regions
found during training, we {\em displace} them towards regions
with better objective scores.
To achieve this, 
during training, HOW clusters data using the procedure of \fig{where}a,
then draws lines between the centroids of  clusters $X,Y$ that are nearest
neighbors.
By convention, we say that the cluster containing $X$ has the  better objective scores. This means we can say that the line is a useful direction for displacing
data; i.e.  away from $Y$ and towards $X$.
Hence,  during testing,  HOW use Pythagoras' theorem
and \fig{where}c to find the  nearest line to every test instance. That test instance is   then displaced towards the better end
of that line (towards  $X$).

An important design decision with HOW 
is how to determine the exact {\em direction} and {\em magnitude} of the
displacement.  Following on from the discussion around   ISSUE~\#1,
we note that a design goal of HOW is
{\em minimal displacements} of project data.
To implement minimal displacement:
\bi
\item
HOW sorts  the independent variables
using the procedure of  \fig{where}d. It then only
displaces the    best $B$\%  of that sort (while leaving the rest untouched).
In  the experiments shown below,  we explore   $B \in \{25,50,100\}\%$.
\item
HOW also  limits displacement
to very small regions in the space of the training data (never more that the separation
of the two closest clusters).  HOW's clusterer divides  data into many   small clusters containing
$\sqrt{N}$ of the population.
HOW   limits displacement to  \mbox{$0 \le F \le c$}, where
$c$ is the distance between the two closest clusters. 
In  the experiments shown below,  we explore
\mbox{$F\in \{0.25, 0.5, 0.75\}$}.
\ei
Turning now to the displacement {\em direction},
our experiments show that it useful to  carefully reflect on exactly where HOW should  push each test instance.
HOW replaces each cluster   with one {\em exemplar} that best
characterizes the data in that cluster. It then displaces the
test data towards that exemplar, as follows.
Let $Z_j$ be an independent variable within the
test instance $Z$. If  $X$ be the exemplar found at training time,
then we displace $Z$ as follows:
\[Z_j =  Z_j + F*(X_j - Z_j)\]
Here, $F$ is the  magnitude controller, discussed above.
The experiments of this work showed that different kinds of data require
different kinds of exemplars.
\bi
\item
For the Seigmund data,
the objective score is the runtime of the test suite of a compiled program.
For this kind of data with numeric objectives,
the exemplar  is the instance in a cluster
with  best objective score.
\item
For the Jureczko data, the objective score
is a boolean that is true if defects have been noted
for that class. 
For this kind of data with discrete objectives,
the data, 
has less differentiation between the objective scores
so we set the exemplar to  
an instance synthesized from the median values of the instances
in that cluster.
\ei

@@

\subsection{How to Assess ``HOW''}
In the experiments shown below,  HOW comments  on how to reduce
software defects by adjusting certain static code parameters of the code\footnote{Such parameters can be manually adjusted by programmers or (for large scale changes) automatically adjusted using code refactoring
tools, as shown by Binkely et al. in 2006~\cite{Binkley2006} and more recently by Mkaouer et al. in 2015~\cite{Mkaouer15}.}. How are we to verify that HOW's recommendations actually work?

Very few industrial or research groups have access
to the kinds of resources using by Bente et al. (evidence: in the size years since the
publication of that work, we know of only one of two similar studies). Hence, it
we wish to verify HOW's plans, then some other framework is required that is less
resource intensive.

We propose using prediction systems (built by data mining) as a verification
tool for planners. In domains with data, from which data miners can build
predictors of adequate performance, then those predictors can assess the value
of the changes proposed by HOW.

As shown in \fig{work}, our proposal divides
project data  into two disjoint sets {\em train} and {\em test}
(so \mbox{{\em train} $\cup$ {\em test} $=\;\emptyset$}).
Next, from the train set, we build both a {\em planner} as well
as a {  predictor}. Our general framework does not   commit to any particular  choice
of { planner} or { predictor} but, for the purposes of this paper, 
our { planner}
will be HOW  and our
our { predictors} will be the Random Forest and Random Forest
Regressor taken from the SciKit
Learn toolkit~\cite{Pedregosa2012}. The Random Forests procedure is explained in \fig{rr}.

\begin{figure}[t!]
\small
\begin{tabular}{|p{.95\linewidth}|}\hline
 Decision tree learners find the attribute whose value's
select for for sections of the data with least variation in the class
variable. Decision tree learners then recurse on each selection~\cite{breiman84}.

A random forest is a set of decision trees, each built
from some random subset of the rows and columns. The conclusions
of the forest are a ``vote'' generated by all the trees~\cite{Breiman2001}.

A standard Random Forest predicts for a discrete class variable while
a Random Forest Regressor generates numeric predictions.\\\hline
\end{tabular}
\caption{Random Forests and Random Forest Regressors.}\label{fig:rr}
\end{figure}

Turning now to the {\em test} data, this is passed to the { predictor}
to generate a baseline {\em before} performance score.
Note that if our { predictors} fail to perform adequately on the test data,
then we cannot trust them to comment on our plans. Accordingly,
if that performance is unsatisfactory, we abort.
Else, we (1)~apply the { planner} to alter the {\em test} data;
then (2)~apply the { predictor} to the altered data {\em test'};
then (3)~return the performance delta.

\begin{figure*}[!t]
\scriptsize
   \begin{center}
   \begin{minipage}{.46\linewidth}
    \begin{tabular}{r@{~}|l@{~}|r@{~}|l@{~}|r@{~}|r@{~}|} \cline{2-6}
   & \multicolumn{5}{c|}{ }\\
